Extreme climatic events with asymmetric tail-dependence affect population-extinction risk

### 2019-07-27 MainText.Rmd
started writing discussion

With global climate change, extreme climatic events (ECEs) appear to be more common than usual. These ECEs 
could pose severe impact on the resistance or resilience of a population or community level through
life-history traits of the species. Therefore predicting the impact of ECEs 
on extinction risk of a population is necessery. We here introduced the potential use of "tail-dependence" 
property of a copula in modelling the extinction risk of an age-structured single patch model. We also 
extended the generalized findings of density-independent metapopulation model from [@Ghosh_copula] 
to density-dependence metapopulation models with this copula approach. For metapopulation models, 
we showed extinction risk not only depended on the nature of tail-dependence of spatially correlated 
noise but was also influenced by the inherent dynamics of the population-model.

@@

<!--stochastic matrix model : age structured-->
To investigate (Q1) from \nameref{Introduction} that how copula structure in noisy 
environment affect the extinction risk of
populations in a single patch age-structured model,
we considered a basic stochastic age-structured model (Eq. \ref{eq_age_str}) 
with two age classes (egg $E$ and adult $A$). We assumed
the population was semelperous, i.e., adult ($A$) went through the
reproductive episode only once before death. Eggs and adults in the 
next generation depend upon the fecundity of adult ($f_a \geq 0$) 
and survival of eggs ($0 \leq s_e \leq 1$) from present generation, respectively.
The environmental stochasticity in the model was embedded in 
the fluctuating vital rates ($f_a$ and $s_e$). We considered the fecundity ($f_a$)
and survival ($s_e$) rates were random draw from a gamma ($\Gamma(k,\theta)$) and
a beta ($\mathcal{B}(\alpha,\beta)$) distribution, respectively. These distributions 
could be generated from iid noises showing specific tail-dependence (left or right tail)
between fecundity and survival rates. Eq. \ref{eq_vitalrates} represent the stochasticity 
in the vital rates. We included the density dependence effect in one of the vital rate,
namely, fecundity ($f_a$). If $\Omega = \infty$ in Eq. \ref{eq_vitalrates}, 
the projection matrix would be independent
of populations from any age-class and the model was then called to be "linear", whereas
its "nonlinear" version had finite $\Omega$.

@@

To test our hypothesis from (Q1), we repeated the above analysis 
considering the vital rates coming from an asymmetric copula with specific 
left (Clayton, Survival Gumbel)
or right (Survival Clayton, Gumbel) tail dependence
or from an extreme tail-depentdent one, 
when vital rates were considered to be affected
by extreme climatic events. A $N \times 2$ matrix, M, was generated
using `BiCopSim` function
from `VineCopula` package 
for each of the left 
(Clayton, Survival Gumbel) and right 
(Survival Clayton, Gumbel) tail-dependent bivariate copulas, 
where $N = numsteps(=25) \times numsims(=10000)$.
We kept the same Kendall's correlation ($\tau = 0.6$) for 
each bivariate copula from one parameter family which used its parameter accordingly
(see `iTau` function from `copula` package). Entries within each clumn of
matrix $M$, were thus 
independent of each others but 
the two columns of $M$ had a specified Kendall's 
correlation ($\tau = 0.6$) and underlying tail-dependence. 
Similarly, for extreme-tail dependent bivariant copula
(either left or right tail) a matrix M was generated from standard iid as described in 
Appendix \ref{SM-extr_tail_dep}.

@@

<!--summary results-->

@@

@@

<!--future study: worthy for PVA, IUCN red list-->

>>>>

With global climate change, extreme climatic events (ECEs) appear to be more common than usual. These ECEs 
could pose severe impact on the resistance or resilience of a population or community level through
life-history traits of the species. Therefore predicting the impact of ECEs 
on extinction risk of a population is necessery. We here introduced the potential use of "tail-dependence" 
property of a copula in modelling the extinction risk of an age-structured single patch model. We also 
extended the generalized findings of density-independent metapopulation model from @Ghosh_copula 
to density-dependence metapopulation models with this copula approach. For metapopulation models, 
we showed extinction risk not only depended on the nature of tail-dependence of spatially correlated 
noise but was also influenced by the inherent dynamics of the population-model.

@@

<!--stochastic matrix model : age structured-->
To investigate (Q1) from \nameref{Introduction} that how copula structure in noisy 
environment affect the extinction risk of
populations in a single patch age-structured model,
we considered a basic stochastic age-structured model (Eq. \ref{eq_age_str}) 
with two age classes (egg $E$ and adult $A$). We assumed
the population was semelperous, i.e., adult ($A$) went through the
reproductive episode only once before death. Eggs and adults in the 
next generation depend upon the fecundity of adult ($f_a \geq 0$) 
and survival of eggs ($0 \leq s_e \leq 1$) from present generation, respectively.
The environmental stochasticity in the model was embedded in 
the fluctuating vital rates ($f_a$ and $s_e$). We considered the fecundity ($f_a$)
and survival ($s_e$) rates were random draw from a gamma ($\Gamma(k,\theta)$) and
a beta ($\mathcal{B}(\alpha,\beta)$) distribution, respectively. These distributions 
were generated from iid noises showing specific tail-dependence (either in left- or in right-tail)
between fecundity and survival rates. Tail dependence emerged if extreme bad years ("catastrophs") 
would suppress both vital rates (hence, $f_a$ and $s_e$ would show a left tail dependent copula), whereas 
extreme good years ("bonanzas") would increase both of them (i.e., $f_a$ and $s_e$ would show 
a right tail dependent copula). Eq. \ref{eq_vitalrates} represent the stochasticity 
in the vital rates. We included the density dependence effect in one of the vital rate,
namely, fecundity ($f_a$). If $\Omega = \infty$ in Eq. \ref{eq_vitalrates}, 
the projection matrix would be independent
of populations from any age-class and the model was then called to be "linear", whereas
its "nonlinear" version had finite $\Omega$.

@@

To test our hypothesis from (Q1), we repeated the above analysis 
considering the vital rates coming from an asymmetric copula with specific 
left (Clayton, Survival Gumbel)
or right (Survival Clayton, Gumbel) tail dependence
or from an extreme tail-depentdent one, 
when vital rates were considered to be affected
by extreme climatic events. A $N \times 2$ matrix, M, was generated
using `BiCopSim` function from `VineCopula` package 
for each of the left (Clayton, Survival Gumbel) and right 
(Survival Clayton, Gumbel) tail-dependent bivariate copulas, 
where $N$ is number of years ($25$) multiplied by number of simulations ($10000$) we used. 
We kept the same Kendall's correlation ($\tau = 0.6$) for 
each bivariate copula from one parameter family which used its parameter accordingly
(see `iTau` function from `copula` package). Entries within each clumn of
matrix $M$, were thus 
independent of each others but 
the two columns of $M$ had a specified Kendall's 
correlation ($\tau = 0.6$) and underlying tail-dependence. 
Similarly, for extreme-tail dependent bivariant copula
(either left or right tail) a matrix M was generated from standard iid as described in 
Appendix \ref{SM-extr_tail_dep}.

@@

<!--summary results:-->
Our results show the potential use of "asymmetric" copula in estimating the impact of 
extreme ecological events on population-extinction risk. We show that through left and right tail 
dependence in vital rates (respectively, spatially correlated environmental noises), years of extreme events 
("catastrophs" and "bonanzas") affect extinction risk of a single-patch age-structured population
(respectively, density dependent metapopulations) differently. Though left tail dependence is 
expected to show higher extinction risk than the one for right tail dependence for age-structured 
and density dependent metapopulation models, it is not always true 
when metapopulation models with dispersal show over-compensatory dynamics. In the over-compensatory regime,
right tail dependent noises result in higher extinction risk than the left tail dependent one. We show 
that in absence of dispersal, the above-said switching in extinction risk measure occurs near the 
critical parameter value where the deterministic model shows transition from 
under-compensatory to over-compensatory regime.

@@

<!--future study: worthy for PVA, IUCN red list
1. Identify which life stage or demographic process as management target?
2. multisite age str. model with dispersal, directional change in spatio-temporal variation for vital rates.
3. Old study (Doak1994, Ferguson1995) showed including correlation btw vital rates for PVA makes a difference, 
   what if we include TD through copula?
4. What if TD with negative correlation?
e.g. some events that will kill adults will allow more successful recruitment of seeds or other propagules.
-->

@@
